---
title: "Shiny Dashboard Building Blocks - Reference Summary"
author: "AMMnet Senegal Workshop"
date: "June 25–27, 2025"
format:
  html:
    theme: minty
    toc: true
    toc-depth: 6
    toc-location: left
    number-sections: false
---

This handout summarizes key Shiny concepts, functions, and patterns you can use when building interactive dashboards.

## Shiny App Structure

Every Shiny app has two main components:

| Component | Purpose | Location |
|----|----|----|
| `ui` | User Interface - controls layout and inputs | Top-level `ui <- ...` |
| `server` | Logic - computations, data processing, outputs | Inside `server <- function(input, output) { ... }` |

The two are linked using shared `inputId` and `outputId` values.

## Inputs and Outputs

### Input Widgets

| Type | Function | When to Use |
|----|----|----|
| Dropdown | `selectInput()` | For long lists (e.g., Province selection) |
| Checkboxes | `checkboxGroupInput()` | For multiple selections |
| Radio Buttons | `radioButtons()` | For short lists |
| Single Checkbox | `checkboxInput()` | For TRUE/FALSE toggles |
| Slider | `sliderInput()` | For numeric ranges |
| Date | `dateInput()` / `dateRangeInput()` | For date selection |

Each input requires:

`inputId = "input_name", label = "Display Text", choices = c(...)`

The current value is accessed using `input$input_name` inside the server.

### Output Types

| Output              | Render Function (Server) | UI Function       |
|---------------------|--------------------------|-------------------|
| Text                | `renderText()`           | `textOutput()`    |
| Table (static)      | `renderTable()`          | `tableOutput()`   |
| Table (interactive) | `renderDT()`             | `DTOutput()`      |
| Plot (static)       | `renderPlot()`           | `plotOutput()`    |
| Plot (interactive)  | `renderPlotly()`         | `plotlyOutput()`  |
| Map                 | `renderLeaflet()`        | `leafletOutput()` |
| Dynamic UI          | `renderUI()`             | `uiOutput()`      |

**Key rule:** every output has one `render*()` function (server) and one `*Output()` function (UI), linked by the same ID.

## Reactivity

Reactivity allows the app to update automatically when inputs change.

-    Use `reactive({ ... })` to create filtered or computed datasets that update when inputs change.

-   Use `render*()` functions to update outputs.

Example reactive filter:

`province_data <- reactive({   req(input$province_select != "")`

`app_data %>% filter(organisation_unit == input$province_select) })`

## Layout Tools

-    `page_sidebar()`: main layout with sidebar + content

-   `layout_sidebar()`: floating sidebar inside a card

-   `card()`: container for plots/tables

-   `value_box()`: summary KPIs

-    `layout_columns()`: arrange elements horizontally

-   `layout_column_wrap()`: responsive wrapping of multiple boxes

Exaple layout:

```{r eval = FALSE}
layout_columns(   
  value_box(title = "Cases", value = textOutput("total_cases")),  
  value_box(title = "Tests", value = textOutput("total_tests")) 
  )
```

## Conditional UI Elements

-   Use `renderUI()` + `uiOutput()` to dynamically generate inputs or outputs.

-   Use `validate()` and `need()` to display user prompts when required inputs are missing.

```{r eval = FALSE}
output$tpr_checkbox_ui <- renderUI({
  if (all(c("malaria_cases", "malaria_tests") %in% input$series_select)) {
    checkboxInput("include_tpr", "Calculate TPR", value = FALSE)
  }
})

```

## Theming and Styling

-    Use `bslib::bs_theme()` to apply consistent color schemes and fonts.

-   Use `input_dark_mode()` to enable light/dark toggle.

-   Use Bootstrap icons via `bsicons::bs_icon()` for value boxes and UI elements.

`theme = bs_theme(version = 5, bootswatch = "minty")`

## Data Export

Allow users to download tables using `downloadHandler()`.

Example:

```{r eval = FALSE}
output$download_data <- downloadHandler(
  filename = function() { "summary.csv" },
  content = function(file) { write.csv(summary_data(), file) }
)
```

## Useful Resources

-   [Shiny Documentation](https://shiny.posit.co/)

-   [Mastering Shiny Book](https://mastering-shiny.org/)

-   [bslib Theming Guide](https://rstudio.github.io/bslib/)
